Type theory in type theory using single substitutions

### 2024-03-05 types_2024/abstract.tex
stuff

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Ambrus Kaposi\inst{1}
\and
Szumi Xie\inst{1}
}

@@

In this abstract we define type theory in a minimalistic way: we aim
for an intrinsic, that is, well-typed and quotiented definition of the
syntax, that is as small as possible in terms of needed operations and
equations for the substitution calculus. We only introduce operations
that are \emph{forced} upon us. We aim for a language with $\Pi$
types, and it turns out that all the essential ingredients are forced
upon us when we want to introduce the usual rules for $\Pi$ types
(abstraction, application, $\beta$ and $\eta$).

@@

Contexts are either empty or extended with a type.

@@

am instantiation operation $\blank[\blank]$ on types, and a weakening

@@

for $\p$, but arbitrary substitutions.

@@

Now that we introduced instantiation of terms, we have to revisit all
of our terms and provide rules on how to instantiate them: first of
all, we need instantiation rules for $\lam$ and

@@

Then we need instantiation rules for variables, we list these by case
distinction on the substitution: weakening of a variable increases the
index by one. When instantiating with lifted substitutions and single

@@

typeckeck the term equations.\vspace{-1.5em}

@@

[^+][\langle\rangle] : A[p^+][\langle\var\,\vz\rangle] = A \hspace{6em}\Pi\eta : t = \lam\,(t[\p]\cdot\var\,\vz)

@@

This concludes all the rules for type theory with $\Pi$: we summarise

@@

single substitution, lifted substitution), we have instantiation rule
for each operation, there are 5 equations describing how instantiation
acts on variables, and 4 equations which describe general properties
of instantiation on types. The rest of the rules are specific to our
single type former $\Pi$: the only extra requirement is that each
operation should come equipped with an instantiation rule ($\Pi[]$,
$\lam[]$, ${\cdot}[]$). Perhaps surprisingly, this is enough to define
the syntax: there is no need for $\Con$ and $\Sub$ to a form a
category, no need for parallel substitutions, empty substitution,
parallel weakenings, or combinations of these.

We can add more type formers...

Show how to derive CwF.

>>>>

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Ambrus Kaposi
\and
Szumi Xie
}

@@

In this abstract we define type theory in a minimalistic way: we aim
for an intrinsic, that is, well-typed and quotiented definition of the
syntax, that is as small as possible in terms of number of operations
and equations for the substitution calculus. We only introduce
operations that are forced upon us. We aim for a language with $\Pi$
types, and it turns out that we have to add all the ingredients of the
substitution calculus while introducing the usual rules for $\Pi$
types (abstraction, application, $\beta$ and $\eta$). When adding more
type formers, we don't need to extend the substitution calculus
anymore.

@@

Contexts are either empty or are built from a context extended with a
type.

@@

an instantiation operation $\blank[\blank]$ on types, and a weakening

@@

for $\p$, but arbitrary substitutions (including lifted ones).

@@

Now that we have instantiation of terms, we need to revisit all
operations producing terms and provide rules on how to instantiate
them: first of all, we need instantiation rules for $\lam$ and

@@

Then we need instantiation rules for variables, we list these for each
possible substitution: weakening of a variable increases the index by
one; when instantiating with lifted substitutions and single

@@

typeckeck the term equations.

@@

[\p^+][\langle\vz\rangle] : A[\p^+][\langle\var\,\vz\rangle] = A \hspace{6em}\Pi\eta : t = \lam\,(t[\p]\cdot\var\,\vz)

@@

This concludes all the rules for type theory with $\Pi$. We summarise

@@

single substitution, lifted substitution), we have 5 equations
describing how instantiation acts on variables, and 4 equations which
describe general properties of instantiation on types. The rest of the
rules are specific to our single type former $\Pi$: the only extra
requirement is that each operation is equipped with an instantiation
rule ($\Pi[]$, $\lam[]$, ${\cdot}[]$). Perhaps surprisingly, this is
enough to define the syntax: there is no need for $\Con$ and $\Sub$ to
a form a category, no need for parallel substitutions, empty
substitution, parallel weakenings, or other combinations of these.

When adding new type formers, we only need the rules for the type
former, and an extra instantiation (naturality) rule for each
operation. For example, a Coquand-universe can be added by $\U :
\Ty\,\Gamma$, $\El:\Tm\,\Gamma\,\U\ra\Ty\,\Gamma$,
$\c:\Ty\,\Gamma\ra\Tm\,\Gamma\,\U$, $\U\beta : \El\,(\c\,A) = A$,
$\U\eta : \c\,(\El\,a) = a$, and three instantiation rules (note that
we need indexing to avoid inconsistency). In \cite{sogat}, we showed
that any second-order generalised algebraic theory (SOGAT) has a
single substitution presentation.

In the syntax (initial model, quotient inductive-inductive type) of
the above theory, all the rules of categories with families (CwF
\cite{DBLP:journals/corr/abs-1904-00827}) are admissible. That is, by
induction on the single substitution syntax, we can define parallel
substitutions (lists of terms) which are composable and form a
category; we can define instantiation by parallel substitutions for
types and terms, these have the usual universal property of
comprehension. The main ingredient for this construction is the notion
of $\alpha$-normal form: a kind of normal form which is still
quotiented by $\Pi\beta$, $\Pi\eta$, but does not include explicit
substitutions. If a type is in $\alpha$-normal form, we know whether
it is $\Pi$, $\U$, or $\El$ of a term. If a term is in $\alpha$-normal
form, we know whether it is a variable, a $\lam$, an application or a
code for a type (note that any function on an $\alpha$-normal has to
respect $\Pi\beta$, $\Pi\eta$, $\U\beta$, $\U\eta$). We prove
$\alpha$-normalisation (every term has a unique $\alpha$-normal form)
and then use induction on $\alpha$-normal forms to define parallel
instantiation and prove its properties.

We formalised the single substitution calculus with an infinite
hierarchy of types closed under $\Pi$ and $\U$ as a quotient
inductive-inductive type in Agda, proved $\alpha$-normalisation for
it, and derived all the rules for the parallel substitution calculus
(CwF equipped with $\Pi$ types and $\U$).

It is clear that the rules for the single substitution calculus are
all derivable from the CwF-rules. The other direction is however not
true: there are more models of the single substitution calculus than
the parallel substitution calculus. The situation is analogous to the
relationship of combinatory logic and lambda calculus: the former has
more models, but the syntaxes are isomorphic
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Our single substitution
calculus is a minimalistic version of B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} (C-systems are their
parallel substitution analogue): B-systems include telescopes and more
equations than our single substitution calculus, making them
equivalent to C-systems.
